\textbf{Differences in COVID-19 vaccination in the province of Ontario across Health Regions and socio-economic strata}

### 2023-06-28 manuscript_v6/main.tex
updating background

\date{}
\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, interior hidden, boxrule=0pt, sharp corners, borderline west={3pt}{0pt}{shadecolor}, breakable, enhanced]}{\end{tcolorbox}}\fi

@@

As of May of 2023, there have been 765 million confirmed cases of
COVID-19 around the world, including 6.8 million
deaths\textsuperscript{\protect\hyperlink{ref-WHO-Covid}{1}}. Although
this disease is no longer categorized as a global health emergency by
the World Health Organization
(WHO)\textsuperscript{\protect\hyperlink{ref-rigby2023}{2}}, there is
ongoing concern due to continued transmission, surges in cases and
deaths due to new
variants\textsuperscript{\protect\hyperlink{ref-un2023}{3}}, and
weaknesses in health systems around the world that could be exploited by
a novel virus or another public health emergency in the
future\textsuperscript{\protect\hyperlink{ref-mackey2021}{4}}.

In particular, a major weakness that has received attention during the
pandemic has been related to inequalities in vaccine uptake. The rapid
development of vaccines against COVID-19 initially brought the hope of a
rapid end to the pandemic due to the start of vaccination campaigns in
certain parts of the world toward the end of
2020\textsuperscript{\protect\hyperlink{ref-thelancet2021}{5}--\protect\hyperlink{ref-tanne2020}{8}})
but inequalities in vaccine uptake made these pharmaceutical
interventions ultimately unable to replicate the success of the smallpox
vaccination program, which was crucial to control this
disease\textsuperscript{\protect\hyperlink{ref-kayser2021}{9}}.

This problematic is a multifaceted issue resulting from a combination of
factors, among which are failed public health
measures\textsuperscript{\protect\hyperlink{ref-li2021}{10}}, inequality
in vaccine access between high- and low-income
countries\textsuperscript{\protect\hyperlink{ref-gerretsen2021}{11},\protect\hyperlink{ref-tamey2022}{12}},
and vaccine
hesitancy\textsuperscript{\protect\hyperlink{ref-nafilyan2021}{13}}.
Furthermore, it is well established that this issue has affected in
particular individuals in certain underrepresented groups (e.g., Black,
Asian, or Indigenous) as well as individuals with socio-economic
disadvantages\textsuperscript{\protect\hyperlink{ref-willis2021}{14}--\protect\hyperlink{ref-hussain2022}{20}}.

Reasons given for this inequality have included medical mistrust due to
systemic medical
racism\textsuperscript{\protect\hyperlink{ref-stoler2021}{16},\protect\hyperlink{ref-mosby2021}{21}},
mistrust in
vaccines\textsuperscript{\protect\hyperlink{ref-willis2021}{14}}, and
the influence of conspiracy
theories\textsuperscript{\protect\hyperlink{ref-mosby2021}{21}--\protect\hyperlink{ref-freeman2020}{23}}.
However, it is important also to consider that vaccination uptake can be
influenced by geographical (spatial) factors. In this regard,
differences in COVID-19 vaccination rates have been associated with
varied regional attitudes towards
vaccination\textsuperscript{\protect\hyperlink{ref-malik2020}{24}},
spatial differences in vaccine access and supply, vaccination location
availability, and lack of prioritization of areas where vulnerable
groups
reside\textsuperscript{\protect\hyperlink{ref-bogoch2022}{7},\protect\hyperlink{ref-nguyen2021}{25}}.
Other studies have also shown heterogeneity in vaccine uptake within
small governmental administrative units such as
counties\textsuperscript{\protect\hyperlink{ref-mollalo2021}{26}--\protect\hyperlink{ref-bhuiyan2022}{29}},
and that accounting for geographical differences in vaccination can help
predict patterns of booster
uptake\textsuperscript{\protect\hyperlink{ref-wood2022}{30}}.

However, such analyses have been carried out mostly in territories
outside of Canada, where available studies have been focused in certain
cities (such as
Toronto\textsuperscript{\protect\hyperlink{ref-choi2021}{31}}, or
Montreal\textsuperscript{\protect\hyperlink{ref-mckinnon2021}{32}}), or
have explored differences at a province-wide
level\textsuperscript{\protect\hyperlink{ref-guay2022}{18}}. Therefore,
there is a need for studies that explore spatial differences in
vaccination within the Canadian territory and that consequently, can
help identify disparities that need to be addressed within specific
areas in each province.

This need is especially important in the case of Ontario, the most
populated province of Canada. Between 2007 and 2019, Ontario managed
healthcare access to its inhabitants using 14 intra-provincial divisions
called the Local Health Integration Networks (LHINs), which aimed to
provide an integrated health system for the province. However, this
approach was complex, bureaucratic, and resulted in excessive
expenditures, disparities in mortality rates, the deterioration of
certain performance indicators such as wait times and hospital
readmissions, fragmented electronic health systems, the decline of
performance indicators, and inequities in health services
access\textsuperscript{\protect\hyperlink{ref-tsasis2012}{33}--\protect\hyperlink{ref-lysyk2016}{37}}.
Therefore, with the intent of better organizing and delivering care in
late 2019 the provincial government eliminated the LHINs and
incorporated the areas covered by them into six larger Health Regions
(North East, North West, Central, Toronto, West, and
East)\textsuperscript{\protect\hyperlink{ref-dong2022}{35}}.

Because of the relatively recent adoption of the Health Region model and
its alignment with the onset of the COVID-19 pandemic, there is a need
to analyze if there are ongoing disparities in health access under this
approach that need to be addressed before they are exploited by a new
disease or public health threat. In this regard, previous research has
highlighted disparities in the level of activity of each Health
Region\textsuperscript{\protect\hyperlink{ref-sethuram2023}{38}}.
Therefore, analyzing differences in vaccination uptake within the Health
Regions and can help identify which socio-demographic groups are the
most vulnerable and what areas of the province deserve special attention
by decision-makers.

@@

We used a logistic regression model to examine the impact of the Health
Regions in vaccination rates while considering the socio-economic
factors and months covered by the survey
(Table~\ref{tbl-descriptive-stats}) and certain interactions (Race and
Health Region and Race and income), as previous studies have shown that
socio-economic factors and their interactions are significant predictors
of intent of vaccination and vaccination
status\textsuperscript{\protect\hyperlink{ref-nguyen2022}{39}--\protect\hyperlink{ref-cnat2022a}{41}}.
Because we identified differences in representativity between the survey
data and the estimates from the Census, we used an iterative
proportional fitting procedure
(\emph{raking})\textsuperscript{\protect\hyperlink{ref-deming1940}{42}}
to correct the data using data from the Census and Health Region
population totals; and fitted the regression model to the uncorrected
and corrected data. Details regarding the correction can be found in the
Appendix. All analyses were conducted in R 4.2.2 using the packages
\texttt{survey}\textsuperscript{\protect\hyperlink{ref-lumley2011}{43}},\texttt{tidyverse}\textsuperscript{\protect\hyperlink{ref-wickham2019}{44}},
\texttt{quarto}\textsuperscript{\protect\hyperlink{ref-quarto}{45}},
\texttt{modelsummary}\textsuperscript{\protect\hyperlink{ref-modelsummary}{46}},
and
\texttt{gtsummary}\textsuperscript{\protect\hyperlink{ref-gtsummary}{47}}.

@@

Our results show that indeed, there were differences in vaccination odds
across Ontario in certain socio-demographic groups. Specifically, those
who identified as Arab/Middle Eastern, Black, Latin American, having
mixed racial or ethnic background, or that belonged to other groups not
explicitly included in the survey (Southeast Asian, Filipino, West
Asian, and minority groups not identified elsewhere) had vaccination
odds that were between a third and a half of that of individuals that
identified as White or Caucasian (Figure~\ref{fig-models}). These
results are consistent with previous studies that have shown lower
vaccination rates in individuals with the same socio-demographic
characteristics\textsuperscript{\protect\hyperlink{ref-guay2022}{18}--\protect\hyperlink{ref-hussain2022}{20},\protect\hyperlink{ref-carter2022}{48}}.

Lower vaccine uptake in the socio-demographic groups indicated above may
be influenced in part, by vaccine hesitancy and refusal, which have been
associated in underrepresented Canadian individuals with concerns on
vaccine safety, effectiveness, and experiences of racial discrimination
in health
settings\textsuperscript{\protect\hyperlink{ref-cnat2022a}{41},\protect\hyperlink{ref-basta2022}{49}--\protect\hyperlink{ref-cnat2023}{51}}.
However, it has been shown that structural barriers also play an
important role in vaccination uptake. In the case of underrepresented
individuals, such barriers include complex scheduling systems, language
barriers, lack of adequate public transportation, and lack of accessible
vaccination
sites\textsuperscript{\protect\hyperlink{ref-njoku2021}{52}}. In this
regard, it is interesting to note that vaccination venues were scarce in
low socio-economic areas that had the highest burden of COVID-19 in
Toronto and other regions of Ontario around the time covered by the
survey\textsuperscript{\protect\hyperlink{ref-bogoch2022}{7},\protect\hyperlink{ref-iveniuk2021}{53}},
and that pharmacies in the Peel region (an area identified as a
``hotspot'' with high numbers of essential workers and multigenerational
households) could not keep up with vaccine
demand\textsuperscript{\protect\hyperlink{ref-gill2022}{54}}. This
suggests that the observed differences are associated with disparities
in vaccine access that were present during the period covered by the
survey.

@@

This result is likely reflects in part the fact that individuals in
underrepresented groups tend to perform occupations that have been
deemed as ``essential'' in the context of the
pandemic\textsuperscript{\protect\hyperlink{ref-hawkins2020}{55},\protect\hyperlink{ref-ct2021}{56}},
which include workers in the areas of grocery stores, gas stations,
warehouses, distribution, and manufacturing, all being occupations for
which an income within the significant brackets identified in the
analysis is to be expected. In Ontario, these workers had priority for
COVID-19
vaccination\textsuperscript{\protect\hyperlink{ref-mishra2021}{57}}; and
there is evidence of interventions by vaccination staff in certain parts
of the province to encourage vaccination uptake by these
individuals\textsuperscript{\protect\hyperlink{ref-gill2022}{54}}. These
facts, combined with evidence of increased trends in vaccination in this
group
elsewhere\textsuperscript{\protect\hyperlink{ref-nguyen2021b}{58}},
suggest that the type of occupation of individuals in underrepresented
groups played an important role in increasing the odds of vaccination.

@@

First, in this case it is useful to analyze the data considering the
LHINs in each Health Region, because most studies in the literature
focused on Ontario use the LHINs as the base of their analyses. The West
Health Region covers the area previously occupied by the Hamilton
Niagara Haldimand Brant, South West, and Waterloo Wellington LHINs,
whereas the East Health Region covers the area of the former Champlain
and Central East LHINs. Previous research has identified health
disparities in these (mostly rural) regions, such as unequal
distribution of primary care providers, increased mortality, and low
pharmacist
availability\textsuperscript{\protect\hyperlink{ref-shah2019}{59}--\protect\hyperlink{ref-timony2022}{61}}.

Furthermore, there is an ongoing challenge for the health system of the
province with regard to personalized healthcare for marginalized
individuals. For example, the West Health Region has only two Aboriginal
Health Access Centres (community-led primary healthcare organizations
focused on First Nations, Métis, and Inuit communities) to provide care
to an estimated 100,000 Indigenous individuals living in the
area\textsuperscript{\protect\hyperlink{ref-ontariohealth}{62}}. Lack of
access to personalized healthcare affects individuals that may mistrust
the traditional healthcare system due to systemic racism or oppression,
which is known to be the case for Indigenous and Black individuals in
Canada, as these rationales have been associated with observed lower
vaccination rates among these
groups\textsuperscript{\protect\hyperlink{ref-smylie2022}{63},\protect\hyperlink{ref-eissa2021}{64}}.
Taken together, this suggests that healthcare disparities specific to
these underrepresented groups in certain parts of the province impacted
vaccination uptake, and highlights the need of investments in the Health
Regions focused on resources, infrastructure, and specially personnel
that can deliver personalized care to marginalized communities, as it
has been shown that such efforts have improved trust in vaccination in
underrepresented groups
elsewhere\textsuperscript{\protect\hyperlink{ref-schafferderoo2020}{65}}.

@@

Additionally, our analysis did not consider the North West and North
East Health Regions, due to the low number of entries from these areas
in the survey (Figure~\ref{fig-map}). Low representation is expected as
these regions as they only account for 5\% of the total population of
Ontario. However, these areas have the highest proportion of Indigenous
inhabitants\textsuperscript{\protect\hyperlink{ref-ontariohealth}{62}}.
In the context of personalized care, there is a need to collect data
that focuses on these Health Regions where additional health disparities
might be present and possibly understudied.

The results in this study are based on self-reported data, where bias
might be present. However, in the context of COVID-19, it has been shown
that good agreement exists between self-reported and documented
vaccination
status\textsuperscript{\protect\hyperlink{ref-stephenson2022}{66}}, we
believe that our data was able to provide a valid sample of vaccination
uptake in the province. This is supported by the statistically
significant higher vaccination odds that were identified for January of
2022 in the model, which are consistent with province-wide trends
reported by Public Health Ontario (which show a 4\% increase between
early December and January, in contrast to a 2.5\% increase between
October and
November\textsuperscript{\protect\hyperlink{ref-ontario-covid}{67}});
however, the short time window constitutes essentially a ``snapshot''
view of the evolution of the disease, and additional data would be
needed to obtain estimates per racial/ethnic group over time across all
Health Regions that can help inform the existence of other health
disparities.

@@

\leavevmode\vadjust pre{\hypertarget{ref-kayser2021}{}}%
\CSLLeftMargin{9. }%
\CSLRightInline{Kayser V, Ramzan I. Vaccines and vaccination: History
and emerging issues. \emph{Human Vaccines {\&}amp\(\mathsemicolon\)
Immunotherapeutics}. 2021;17(12):5255-5268.
doi:\href{https://doi.org/10.1080/21645515.2021.1977057}{10.1080/21645515.2021.1977057}}

@@

\leavevmode\vadjust pre{\hypertarget{ref-gerretsen2021}{}}%
\CSLLeftMargin{11. }%
\CSLRightInline{Gerretsen P, Kim J, Caravaggio F, et al. Individual
determinants of {COVID}-19 vaccine hesitancy. Inbaraj LR, ed.
\emph{{PLOS} {ONE}}. 2021;16(11):e0258462.
doi:\href{https://doi.org/10.1371/journal.pone.0258462}{10.1371/journal.pone.0258462}}

\leavevmode\vadjust pre{\hypertarget{ref-tamey2022}{}}%
\CSLLeftMargin{12. }%
\CSLRightInline{Yamey G, Garcia P, Hassan F, et al. It is not too late
to achieve global covid-19 vaccine equity. \emph{{BMJ}}. Published
online March 2022:e070650.
doi:\href{https://doi.org/10.1136/bmj-2022-070650}{10.1136/bmj-2022-070650}}

\leavevmode\vadjust pre{\hypertarget{ref-nafilyan2021}{}}%
\CSLLeftMargin{13. }%
\CSLRightInline{Nafilyan V, Dolby T, Razieh C, et al. Sociodemographic
inequality in {COVID}-19 vaccination coverage among elderly adults in
england: A national linked data study. \emph{{BMJ} Open}.
2021;11(7):e053402.
doi:\href{https://doi.org/10.1136/bmjopen-2021-053402}{10.1136/bmjopen-2021-053402}}

\leavevmode\vadjust pre{\hypertarget{ref-willis2021}{}}%
\CSLLeftMargin{14. }%
\CSLRightInline{Willis DE, Andersen JA, Bryant-Moore K, et al.
{COVID}-19 vaccine hesitancy: Race/ethnicity, trust, and fear.
\emph{Clinical and Translational Science}. 2021;14(6):2200-2207.
doi:\href{https://doi.org/10.1111/cts.13077}{10.1111/cts.13077}}

\leavevmode\vadjust pre{\hypertarget{ref-skirrow2022}{}}%
\CSLLeftMargin{15. }%
\CSLRightInline{Skirrow H, Barnett S, Bell S, et al. Women's views on
accepting {COVID}-19 vaccination during and after pregnancy, and for
their babies: A multi-methods study in the {UK}. \emph{{BMC} Pregnancy
and Childbirth}. 2022;22(1).
doi:\href{https://doi.org/10.1186/s12884-021-04321-3}{10.1186/s12884-021-04321-3}}

\leavevmode\vadjust pre{\hypertarget{ref-stoler2021}{}}%
\CSLLeftMargin{16. }%
\CSLRightInline{Stoler J, Enders AM, Klofstad CA, Uscinski JE. The
limits of medical trust in mitigating {COVID}-19 vaccine hesitancy among
black americans. \emph{Journal of General Internal Medicine}.
2021;36(11):3629-3631.
doi:\href{https://doi.org/10.1007/s11606-021-06743-3}{10.1007/s11606-021-06743-3}}

\leavevmode\vadjust pre{\hypertarget{ref-khubchandani2021}{}}%
\CSLLeftMargin{17. }%
\CSLRightInline{Khubchandani J, Sharma S, Price JH, Wiblishauser MJ,
Sharma M, Webb FJ. {COVID}-19 vaccination hesitancy in the united
states: A rapid national assessment. \emph{Journal of Community Health}.
2021;46(2):270-277.
doi:\href{https://doi.org/10.1007/s10900-020-00958-x}{10.1007/s10900-020-00958-x}}

>>>>

\date{}
\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, enhanced, boxrule=0pt, interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, frame hidden]}{\end{tcolorbox}}\fi

@@

As of May of 2023, there have been 765 million confirmed cases of
COVID-19 around the world, including 6.8 million
deaths\textsuperscript{\protect\hyperlink{ref-WHO-Covid}{1}}. Although
this disease is no longer categorized as a global health emergency by
the World Health Organization
(WHO)\textsuperscript{\protect\hyperlink{ref-rigby2023}{2}}, there is
ongoing concern due to continued transmission, the possibility of surges
in cases and deaths due to new virus
variants\textsuperscript{\protect\hyperlink{ref-un2023}{3}}, and ongoing
issues in health systems around the world that could be exploited by a
novel virus or another public health emergency in the
future\textsuperscript{\protect\hyperlink{ref-mackey2021}{4}}.

In particular, the pandemic showed multiple challenges with regard to
vaccination. The rapid development of vaccines against COVID-19
initially brought the hope of a rapid end to the pandemic as vaccination
campaigns in certain parts of the world started as early as December of
2020\textsuperscript{\protect\hyperlink{ref-thelancet2021}{5}--\protect\hyperlink{ref-tanne2020}{8}}.
Although it has been estimated that so far COVID-19 vaccines have been
able to prevent millions of deaths
worldwide\textsuperscript{\protect\hyperlink{ref-watson2022}{9}}, their
implementation has faced multiple challenges across the world. These
challenges included the rapid emergence of new virus
variants\textsuperscript{\protect\hyperlink{ref-li2021}{10}}, the waning
of vaccine
protection\textsuperscript{\protect\hyperlink{ref-dolgin2021}{11}},
inequality in vaccine access between high-income and low-income
countries\textsuperscript{\protect\hyperlink{ref-gerretsen2021}{12},\protect\hyperlink{ref-tamey2022}{13}},
vaccine
hesitancy\textsuperscript{\protect\hyperlink{ref-nafilyan2021}{14}}, and
differences in vaccine uptake across the
population\textsuperscript{\protect\hyperlink{ref-willis2021}{15},\protect\hyperlink{ref-skirrow2022}{16}}.
In particular, it is well established that lower vaccination uptake has
been observed in individuals within certain underrepresented groups
(e.g., Black, Asian, or Indigenous) as well as individuals with
socio-economic
disadvantages\textsuperscript{\protect\hyperlink{ref-willis2021}{15}--\protect\hyperlink{ref-hussain2022}{21}}.

Reasons given for this inequality in vaccination uptake have included
medical mistrust due to systemic medical
racism\textsuperscript{\protect\hyperlink{ref-stoler2021}{17},\protect\hyperlink{ref-mosby2021}{22}},
mistrust in
vaccines\textsuperscript{\protect\hyperlink{ref-willis2021}{15}}, and
the influence of misinformation and
disinformation\textsuperscript{\protect\hyperlink{ref-mosby2021}{22}--\protect\hyperlink{ref-freeman2020}{24}}.
However, it is important also to consider that vaccination uptake can be
influenced by geographical (spatial) factors. In this regard,
differences in COVID-19 vaccination rates have been associated with
varied regional attitudes towards
vaccination\textsuperscript{\protect\hyperlink{ref-malik2020}{25}}, lack
of vaccine access in areas where underrepresented groups
reside\textsuperscript{\protect\hyperlink{ref-bogoch2022}{7}}, as well
as an absence of prioritization of areas inhabited by vulnerable
groups\textsuperscript{\protect\hyperlink{ref-nguyen2021}{26}}. Other
studies have also shown heterogeneity in vaccine uptake within small
governmental administrative units such as
counties\textsuperscript{\protect\hyperlink{ref-mollalo2021}{27}--\protect\hyperlink{ref-bhuiyan2022}{30}},
indicating that accounting for geographical differences in vaccination
can help predict patterns of booster
uptake\textsuperscript{\protect\hyperlink{ref-wood2022}{31}}.

However, such analyses have been carried out mostly in territories
outside of Canada, where available studies have been focused in certain
cities (such as
Toronto\textsuperscript{\protect\hyperlink{ref-choi2021}{32}}, or
Montreal\textsuperscript{\protect\hyperlink{ref-mckinnon2021}{33}}), or
have explored differences at a province-wide
level\textsuperscript{\protect\hyperlink{ref-guay2022}{19}}. Therefore,
there is a need for studies that explore spatial differences in
vaccination within the Canadian territory and that consequently, can
help identify disparities that need to be addressed within specific
areas in each province.

This need is especially important in the case of Ontario, the most
populated province of Canada. Between 2007 and 2019, Ontario managed
healthcare access to its inhabitants using 14 intra-provincial divisions
called the Local Health Integration Networks (LHINs), which aimed to
provide an integrated health system for the province. However, this
approach was complex, bureaucratic, and resulted in excessive
expenditures, disparities in mortality rates, the deterioration of
certain performance indicators such as wait times and hospital
readmissions, fragmented electronic health systems, and inequities in
health services
access\textsuperscript{\protect\hyperlink{ref-tsasis2012}{34}--\protect\hyperlink{ref-lysyk2016}{38}}.
Therefore, with the intent of better organizing and delivering care in
late 2019 the provincial government eliminated the LHINs and
incorporated the areas covered by them into six larger Health Regions
(North East, North West, Central, Toronto, West, and East), which are
managed by the new government agency Ontario Health
(OH)\textsuperscript{\protect\hyperlink{ref-dong2022}{36}}.

On the other hand, public health in Ontario is administered by Public
Health Ontario (PHO), a government agency established in late 2007 and
currently composed of 34 Public Health Units (PHUs) that cover the
entire geography of the province. Although these PHUS were commissioned
with leading the distribution of the COVID-19 vaccine within the areas
they covered\textsuperscript{\protect\hyperlink{ref-rotenberg2021}{39}},
there was significant overlap between PHO and OH in the actual vaccine
rollout, and in many instances vaccines were administered in vaccination
clinics ran by members of OH working alone or in conjunction with
members of
PHO\textsuperscript{\protect\hyperlink{ref-ashcroft2023}{40}}.

\begin{itemize}
\tightlist
\item
  add about intersection of Health Regions and PHUs. How vaccination
  presented a challenge as vaccine was ultimately administered in
  facilites managed by OHT?
\end{itemize}

https://www.mcmasterforum.org/docs/default-source/product-documents/rapid-responses/examining-the-intersections-between-ontario-health-teams-and-public-health.pdf?sfvrsn=69caa8c9\_11

Because of the relatively recent adoption of the Health Region model and
its alignment with the onset of the COVID-19 pandemic, there is a need
to analyze if there are ongoing disparities in health access under this
approach that need to be addressed before they are exploited by a new
disease or public health threat. In this regard, previous research has
highlighted disparities in the level of activity of each Health
Region\textsuperscript{\protect\hyperlink{ref-sethuram2023}{41}}.
Therefore, analyzing differences in vaccination uptake within the Health
Regions and can help identify which socio-demographic groups are the
most vulnerable and what areas of the province deserve special attention
by decision-makers.

@@

We used a logistic regression model to examine the impact of the Health
Regions in vaccination rates while considering the socio-economic
factors and months covered by the survey
(Table~\ref{tbl-descriptive-stats}) and certain interactions (Race and
Health Region and Race and income), as previous studies have shown that
socio-economic factors and their interactions are significant predictors
of intent of vaccination and vaccination
status\textsuperscript{\protect\hyperlink{ref-nguyen2022}{42}--\protect\hyperlink{ref-cnat2022a}{44}}.
Because we identified differences in representativity between the survey
data and the estimates from the Census, we used an iterative
proportional fitting procedure
(\emph{raking})\textsuperscript{\protect\hyperlink{ref-deming1940}{45}}
to correct the data using data from the Census and Health Region
population totals; and fitted the regression model to the uncorrected
and corrected data. Details regarding the correction can be found in the
Appendix. All analyses were conducted in R 4.2.2 using the packages
\texttt{survey}\textsuperscript{\protect\hyperlink{ref-lumley2011}{46}},\texttt{tidyverse}\textsuperscript{\protect\hyperlink{ref-wickham2019}{47}},
\texttt{quarto}\textsuperscript{\protect\hyperlink{ref-quarto}{48}},
\texttt{modelsummary}\textsuperscript{\protect\hyperlink{ref-modelsummary}{49}},
and
\texttt{gtsummary}\textsuperscript{\protect\hyperlink{ref-gtsummary}{50}}.

@@

Our results show that indeed, there were differences in vaccination odds
across Ontario in certain socio-demographic groups. Specifically, those
who identified as Arab/Middle Eastern, Black, Latin American, having
mixed racial or ethnic background, or that belonged to other groups not
explicitly included in the survey (Southeast Asian, Filipino, West
Asian, and minority groups not identified elsewhere) had vaccination
odds that were between a third and a half of that of individuals that
identified as White or Caucasian (Figure~\ref{fig-models}). These
results are consistent with previous studies that have shown lower
vaccination rates in individuals with the same socio-demographic
characteristics\textsuperscript{\protect\hyperlink{ref-guay2022}{19}--\protect\hyperlink{ref-hussain2022}{21},\protect\hyperlink{ref-carter2022}{51}}.

Lower vaccine uptake in the socio-demographic groups indicated above may
be influenced in part, by vaccine hesitancy and refusal, which have been
associated in underrepresented Canadian individuals with concerns on
vaccine safety, effectiveness, and experiences of racial discrimination
in health
settings\textsuperscript{\protect\hyperlink{ref-cnat2022a}{44},\protect\hyperlink{ref-basta2022}{52}--\protect\hyperlink{ref-cnat2023}{54}}.
However, it has been shown that structural barriers also play an
important role in vaccination uptake. In the case of underrepresented
individuals, such barriers include complex scheduling systems, language
barriers, lack of adequate public transportation, and lack of accessible
vaccination
sites\textsuperscript{\protect\hyperlink{ref-njoku2021}{55}}. In this
regard, it is interesting to note that vaccination venues were scarce in
low socio-economic areas that had the highest burden of COVID-19 in
Toronto and other regions of Ontario around the time covered by the
survey\textsuperscript{\protect\hyperlink{ref-bogoch2022}{7},\protect\hyperlink{ref-iveniuk2021}{56}},
and that pharmacies in the Peel region (an area identified as a
``hotspot'' with high numbers of essential workers and multigenerational
households) could not keep up with vaccine
demand\textsuperscript{\protect\hyperlink{ref-gill2022}{57}}. This
suggests that the observed differences are associated with disparities
in vaccine access that were present during the period covered by the
survey.

@@

This result is likely reflects in part the fact that individuals in
underrepresented groups tend to perform occupations that have been
deemed as ``essential'' in the context of the
pandemic\textsuperscript{\protect\hyperlink{ref-hawkins2020}{58},\protect\hyperlink{ref-ct2021}{59}},
which include workers in the areas of grocery stores, gas stations,
warehouses, distribution, and manufacturing, all being occupations for
which an income within the significant brackets identified in the
analysis is to be expected. In Ontario, these workers had priority for
COVID-19
vaccination\textsuperscript{\protect\hyperlink{ref-mishra2021}{60}}; and
there is evidence of interventions by vaccination staff in certain parts
of the province to encourage vaccination uptake by these
individuals\textsuperscript{\protect\hyperlink{ref-gill2022}{57}}. These
facts, combined with evidence of increased trends in vaccination in this
group
elsewhere\textsuperscript{\protect\hyperlink{ref-nguyen2021b}{61}},
suggest that the type of occupation of individuals in underrepresented
groups played an important role in increasing the odds of vaccination.

@@

First, in this case it is useful to analyze the data considering the
LHINs in each Health Region, because most studies in the literature
focused on Ontario use the LHINs as the base of their analyses. The West
Health Region covers the area previously occupied by the Hamilton
Niagara Haldimand Brant, South West, and Waterloo Wellington LHINs,
whereas the East Health Region covers the area of the former Champlain
and Central East LHINs. Previous research has identified health
disparities in these (mostly rural) regions, such as unequal
distribution of primary care providers, increased mortality, and low
pharmacist
availability\textsuperscript{\protect\hyperlink{ref-shah2019}{62}--\protect\hyperlink{ref-timony2022}{64}}.

Furthermore, there is an ongoing challenge for the health system of the
province with regard to personalized healthcare for marginalized
individuals. For example, the West Health Region has only two Aboriginal
Health Access Centres (community-led primary healthcare organizations
focused on First Nations, Métis, and Inuit communities) to provide care
to an estimated 100,000 Indigenous individuals living in the
area\textsuperscript{\protect\hyperlink{ref-ontariohealth}{65}}. Lack of
access to personalized healthcare affects individuals that may mistrust
the traditional healthcare system due to systemic racism or oppression,
which is known to be the case for Indigenous and Black individuals in
Canada, as these rationales have been associated with observed lower
vaccination rates among these
groups\textsuperscript{\protect\hyperlink{ref-smylie2022}{66},\protect\hyperlink{ref-eissa2021}{67}}.
Taken together, this suggests that healthcare disparities specific to
these underrepresented groups in certain parts of the province impacted
vaccination uptake, and highlights the need of investments in the Health
Regions focused on resources, infrastructure, and specially personnel
that can deliver personalized care to marginalized communities, as it
has been shown that such efforts have improved trust in vaccination in
underrepresented groups
elsewhere\textsuperscript{\protect\hyperlink{ref-schafferderoo2020}{68}}.

@@

Additionally, our analysis did not consider the North West and North
East Health Regions, due to the low number of entries from these areas
in the survey (Figure~\ref{fig-map}). Low representation is expected as
these regions as they only account for 5\% of the total population of
Ontario. However, these areas have the highest proportion of Indigenous
inhabitants\textsuperscript{\protect\hyperlink{ref-ontariohealth}{65}}.
In the context of personalized care, there is a need to collect data
that focuses on these Health Regions where additional health disparities
might be present and possibly understudied.

The results in this study are based on self-reported data, where bias
might be present. However, in the context of COVID-19, it has been shown
that good agreement exists between self-reported and documented
vaccination
status\textsuperscript{\protect\hyperlink{ref-stephenson2022}{69}}, we
believe that our data was able to provide a valid sample of vaccination
uptake in the province. This is supported by the statistically
significant higher vaccination odds that were identified for January of
2022 in the model, which are consistent with province-wide trends
reported by Public Health Ontario (which show a 4\% increase between
early December and January, in contrast to a 2.5\% increase between
October and
November\textsuperscript{\protect\hyperlink{ref-ontario-covid}{70}});
however, the short time window constitutes essentially a ``snapshot''
view of the evolution of the disease, and additional data would be
needed to obtain estimates per racial/ethnic group over time across all
Health Regions that can help inform the existence of other health
disparities.

@@

\leavevmode\vadjust pre{\hypertarget{ref-watson2022}{}}%
\CSLLeftMargin{9. }%
\CSLRightInline{Watson OJ, Barnsley G, Toor J, Hogan AB, Winskill P,
Ghani AC. Global impact of the first year of {COVID}-19 vaccination: A
mathematical modelling study. \emph{The Lancet Infectious Diseases}.
2022;22(9):1293-1302.
doi:\href{https://doi.org/10.1016/s1473-3099(22)00320-6}{10.1016/s1473-3099(22)00320-6}}

@@

\leavevmode\vadjust pre{\hypertarget{ref-dolgin2021}{}}%
\CSLLeftMargin{11. }%
\CSLRightInline{Dolgin E. {COVID} vaccine immunity is waning
{\textemdash} how much does that matter? \emph{Nature}.
2021;597(7878):606-607.
doi:\href{https://doi.org/10.1038/d41586-021-02532-4}{10.1038/d41586-021-02532-4}}

\leavevmode\vadjust pre{\hypertarget{ref-gerretsen2021}{}}%
\CSLLeftMargin{12. }%
\CSLRightInline{Gerretsen P, Kim J, Caravaggio F, et al. Individual
determinants of {COVID}-19 vaccine hesitancy. Inbaraj LR, ed.
\emph{{PLOS} {ONE}}. 2021;16(11):e0258462.
doi:\href{https://doi.org/10.1371/journal.pone.0258462}{10.1371/journal.pone.0258462}}

\leavevmode\vadjust pre{\hypertarget{ref-tamey2022}{}}%
\CSLLeftMargin{13. }%
\CSLRightInline{Yamey G, Garcia P, Hassan F, et al. It is not too late
to achieve global covid-19 vaccine equity. \emph{{BMJ}}. Published
online March 2022:e070650.
doi:\href{https://doi.org/10.1136/bmj-2022-070650}{10.1136/bmj-2022-070650}}

\leavevmode\vadjust pre{\hypertarget{ref-nafilyan2021}{}}%
\CSLLeftMargin{14. }%
\CSLRightInline{Nafilyan V, Dolby T, Razieh C, et al. Sociodemographic
inequality in {COVID}-19 vaccination coverage among elderly adults in
england: A national linked data study. \emph{{BMJ} Open}.
2021;11(7):e053402.
doi:\href{https://doi.org/10.1136/bmjopen-2021-053402}{10.1136/bmjopen-2021-053402}}

\leavevmode\vadjust pre{\hypertarget{ref-willis2021}{}}%
\CSLLeftMargin{15. }%
\CSLRightInline{Willis DE, Andersen JA, Bryant-Moore K, et al.
{COVID}-19 vaccine hesitancy: Race/ethnicity, trust, and fear.
\emph{Clinical and Translational Science}. 2021;14(6):2200-2207.
doi:\href{https://doi.org/10.1111/cts.13077}{10.1111/cts.13077}}

\leavevmode\vadjust pre{\hypertarget{ref-skirrow2022}{}}%
\CSLLeftMargin{16. }%
\CSLRightInline{Skirrow H, Barnett S, Bell S, et al. Women's views on
accepting {COVID}-19 vaccination during and after pregnancy, and for
their babies: A multi-methods study in the {UK}. \emph{{BMC} Pregnancy
and Childbirth}. 2022;22(1).
doi:\href{https://doi.org/10.1186/s12884-021-04321-3}{10.1186/s12884-021-04321-3}}

\leavevmode\vadjust pre{\hypertarget{ref-stoler2021}{}}%
\CSLLeftMargin{17. }%
\CSLRightInline{Stoler J, Enders AM, Klofstad CA, Uscinski JE. The
limits of medical trust in mitigating {COVID}-19 vaccine hesitancy among
black americans. \emph{Journal of General Internal Medicine}.
2021;36(11):3629-3631.
doi:\href{https://doi.org/10.1007/s11606-021-06743-3}{10.1007/s11606-021-06743-3}}

\leavevmode\vadjust pre{\hypertarget{ref-khubchandani2021}{}}%
\CSLLeftMargin{18. }%
\CSLRightInline{Khubchandani J, Sharma S, Price JH, Wiblishauser MJ,
Sharma M, Webb FJ. {COVID}-19 vaccination hesitancy in the united
states: A rapid national assessment. \emph{Journal of Community Health}.
2021;46(2):270-277.
doi:\href{https://doi.org/10.1007/s10900-020-00958-x}{10.1007/s10900-020-00958-x}}
